Constructing a universe for the setoid model

### 2021-01-22 paper.tex
remove mentions of formalisation being unfinished; add back the last paragraph

This section outlines a way to encode the inductive-inductive universe of
setoids from the previous section without assuming arbitrary inductive-inductive
definitions in the metatheory.

@@

Following \cite{ii}, we distinguish two versions of the eliminator:
the simple and the general one. Note that this is orthogonal to the
distinction between nondependent and dependent eliminators, from which
we only consider the latter. The motives for the simple eliminator are
$C':\textsf{Con}\to\mType$,
$T':(\GG:\textsf{Con})(A:\textsf{Ty}\,\GG)\to \mType$ and the
eliminators themselves have the following signatures:
\begin{alignat*}{10}
  & \textsf{elim}_{\textsf{Con}}' && : (\GG : \textsf{Con}) \to C'\ \GG \hspace{3.3em} \textsf{elim}_{\textsf{Ty}}' && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T'\ \GG\ A \hphantom{(\textsf{elim}_{\textsf{Con}}\ \GG)}
\end{alignat*}
In the case of the general eliminator, the motive for $\textsf{Ty}$
depends on the motive for $\textsf{Con}$, making the two eliminators
\emph{recursive-recursive} functions. For motives $C :
\textsf{Con} \to \mType$ and $T : (\GG : \textsf{Con}) \to
\textsf{Ty}\ \GG \to C\ \GG \to \mType$ the signatures are:
\begin{alignat*}{10}
  & \textsf{elim}_{\textsf{Con}} && : (\GG : \textsf{Con}) \to C\ \GG \hphantom{'}\hspace{3.3em} \textsf{elim}_{\textsf{Ty}} && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
\end{alignat*}

@@

Luckily, it would seem the \emph{simple} elimination principle is sufficient for
our purposes: careful inspection of the code suggests that all functions
described in Section \ref{ii-universe} can be defined using simple mutual
pattern matching without recursion-recursion. The simple eliminator can be
defined by pattern matching on the untyped codes, and does not require
extensionality or any extra principles beyond strong transport. Note that
producing a full formalization that uses exclusively simple eliminators appears
to be extremely time-consuming; we therefore leave this task to future work.

@@

We have described the construction of a universe of setoids in the setoid model
of type theory; this is given in several steps, first as an inductive-recursive
definition, then as an inductive-inductive definition, and finally as an
inductive type. Every encoding is obtained from the previous by adapting known
data type transformation methods in a novel way that accounts for the
peculiarities of our construction. In \cite{mpc19} we present rules
for SetTT, clearly these rules need to be extended by the rules for a
universe reflecting the semantics presented here.
%
Most of the mathematical contents of this paper have been formalized in the
proof-assistant Agda (see \cite{agda-code}). We aim to complete the
formalization in the near future, in particular to obtain a machine-checked
proof that our inductive construction of the setoid universe does not rely on
recursion-recursion.

@@

% Groupoids can be regarded as generalized setoids. In the future we would like to
% design a type theory internalizing the groupoid model of type theory
% \cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
% model. A further question is whether such ``groupoid type theory'' can be
% justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
% itself as the target theory.

>>>>

This section encodes the inductive-inductive universe of setoids from
the previous section without assuming arbitrary inductive-inductive
definitions in the metatheory.

@@

Following \cite{ii}, we distinguish two versions of the eliminator:
the \emph{simple} and the \emph{general} one. Note that this is orthogonal to the
distinction between nondependent and dependent eliminators, from which
we only consider the latter. The motives for the simple eliminator are
$C':\textsf{Con}\to\mType$,
$T':(\GG:\textsf{Con})(A:\textsf{Ty}\,\GG)\to \mType$ and the
eliminators themselves have the following signatures:
\begin{alignat*}{10}
  & \textsf{elim}_{\textsf{Con}}' && : (\GG : \textsf{Con}) \to C'\ \GG \hspace{3.3em} \textsf{elim}_{\textsf{Ty}}' && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T'\ \GG\ A \hphantom{(\textsf{elim}_{\textsf{Con}}\ \GG)}
\end{alignat*}
In the case of the general eliminator, the motive for $\textsf{Ty}$
depends on the motive for $\textsf{Con}$, making the two eliminators
\emph{recursive-recursive} functions. For motives $C :
\textsf{Con} \to \mType$ and $T : (\GG : \textsf{Con}) \to
\textsf{Ty}\ \GG \to C\ \GG \to \mType$ the signatures are:
\begin{alignat*}{10}
  & \textsf{elim}_{\textsf{Con}} && : (\GG : \textsf{Con}) \to C\ \GG \hphantom{'}\hspace{3.3em} \textsf{elim}_{\textsf{Ty}} && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
\end{alignat*}

@@

Luckily, the \emph{simple} elimination principle is sufficient for our
purposes: all functions described in Section \ref{ii-universe} can be
defined just using the simple eliminator without
recursion-recursion. The simple eliminator itself can be defined by
pattern matching on the untyped codes, and does not require
extensionality or any extra principles beyond strong transport.

@@

We have described the construction of a universe of setoids in the setoid model
of type theory; this is given in several steps, first as an inductive-recursive
definition, then as an inductive-inductive definition, and finally as an
inductive type. Every encoding is obtained from the previous by adapting known
data type transformation methods in a novel way that accounts for the
peculiarities of our construction. In \cite{mpc19} we present rules
for SetTT, clearly these rules need to be extended by the rules for a
universe reflecting the semantics presented here.
%
% Most of the mathematical contents of this paper have been formalized in the
% proof-assistant Agda (see \cite{agda-code}). We aim to complete the
% formalization in the near future, in particular to obtain a machine-checked
% proof that our inductive construction of the setoid universe does not rely on
% recursion-recursion.

@@

Groupoids can be regarded as generalized setoids. In the future we would like to
design a type theory internalizing the groupoid model of type theory
\cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
model. A further question is whether such ``groupoid type theory'' can be
justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
itself as the target theory.
